---
title: "Understanding Maternal Mortality in America"
author: "Malika Dosso"
date: "2024-10-21"
output: html_document
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, comment = NA)
```


\Research Findings: US Maternal Mortality Analysis (2019-2024)

## 1. Overall Statistical Overview
* Complete Study Period Average: 22.97 deaths per 100,000 live births
* Range Analysis:
  - Absolute Minimum: 17.3 deaths per 100,000 live births (Pre-COVID)
  - Absolute Maximum: 33.8 deaths per 100,000 live births (February 2022)
  - Total Span: Initial rate 17.4 to final rate 19.6
* Volatility Measures:
  - Pre-COVID SD: 1.04
  - COVID Era SD: 4.67
  - Post-Dobbs SD: Variable (showing declining trend)

## 2. Pre-COVID Period (2019-March 2020) 
Baseline Period Characteristics:
* Duration: 14 months
* Statistical Measures:
  - Mean Rate: 18.9 deaths per 100,000 live births
  - Standard Deviation: 1.04 (most stable period)
  - Range: 17.3-20.1
* Healthcare Metrics:
  - Average Monthly Deaths: 712
  - Average Monthly Births: 3,761,836
* Trend Analysis:
  - Steady increase of 13.8% over period
  - Most consistent period with predictable patterns
  - Established baseline for comparison

## 3. COVID Era (March 2020-June 2022)
Peak Period Analysis:
* Duration: 27 months
* Critical Statistics:
  - Mean Rate: 26.4 deaths per 100,000 live births
  - Peak Rate: 33.8 (February 2022)
  - Standard Deviation: 4.67 (highest volatility)
* Impact Metrics: 
  - Total Increase: 57.7%
  - Average Monthly Deaths: 962 (250 more than pre-COVID)
  - Average Monthly Births: 3,648,065 (decline of 113,771)
* Pattern Analysis:
  - Consistent upward trend
  - Highest volatility in rates
  - Significant monthly variations

## 4. Post-Dobbs Period (June 2022-Present)
Transition and Recovery Phase:
* Duration: 25 months
* Key Metrics:
  - Mean Rate: 21.6 deaths per 100,000 live births
  - Initial Rate: 32.1
  - Final Rate: 19.6
  - Total Decline: 38.9%
* Volume Statistics:
  - Average Monthly Deaths: 786
  - Average Monthly Births: 3,639,959
* Trend Analysis:
  - Consistent downward trend
  - Moving toward pre-COVID levels
  - Stabilizing patterns emerging

## 5. Racial Disparity Analysis
Detailed Breakdown by Race/Ethnicity:

American Indian/Alaska Native:
* Highest Overall Rates:
  - COVID Era Mean: 98.2
  - Post-Dobbs Mean: 86.5
  - Disparity Ratio: 3.90x higher than White rate

Black, Non-Hispanic:
* Consistently Second Highest:
  - COVID Era Mean: 59.0
  - Post-Dobbs Mean: 51.3
  - Disparity Ratio: 2.78x higher than White rate

White, Non-Hispanic (Reference Group):
* Baseline Metrics:
  - COVID Era Mean: 21.4
  - Post-Dobbs Mean: 17.6
  - Used as reference for disparity calculations

Hispanic:
* Variable Patterns:
  - COVID Era Mean: 21.0
  - Post-Dobbs Mean: 15.9
  - Disparity Ratio: 0.973x (near parity with White rate)

Asian, Non-Hispanic:
* Consistently Lowest Rates:
  - COVID Era Mean: 14.2
  - Post-Dobbs Mean: 12.8
  - Disparity Ratio: 0.666x lower than White rate

## 6. Annual Progression Analysis
Yearly Rate Changes:
* 2019: 18.8 (Baseline Year)
  - Stable patterns
  - Pre-pandemic normal
* 2020: 21.4
  - Increase: 14.2%
  - Initial COVID impact
* 2021: 27.5
  - Increase: 28.1%
  - Peak COVID impact
* 2022: 29.5
  - Increase: 7.34%
  - Transition year (COVID/Dobbs)
* 2023: 19.7
  - Decrease: 33.3%
  - Major recovery period
* 2024: 19.0
  - Decrease: 3.09%
  - Continuing stabilization

## 7. Birth Rate Trends
Longitudinal Birth Analysis:
* Pre-COVID Average: 3,761,836
  - Highest birth rates
  - Stable patterns
* COVID Era Average: 3,648,065
  - Decline of 113,771 births
  - Consistent decrease
* Post-Dobbs Average: 3,639,959
  - Further slight decline
  - Minimal change from COVID era
  
  
```{r}
library(tidyverse)
library(lubridate)
```

```{r}
VSRR_MM <- read.csv("~/Downloads/VSRR_Provisional_Maternal_Death_Counts_and_Rates.csv")
```

```{r}
head(VSRR_MM)
str(VSRR_MM)
summary(VSRR_MM)
names(VSRR_MM)
colSums(is.na(VSRR_MM))
```

```{r}
# Cleaning and formatting the data
VSRR_MM_clean <- VSRR_MM %>%
  # Converting character numbers to numeric
  mutate(
    # Removing commas and converting to numeric
    Maternal.Deaths = as.numeric(gsub(",", "", Maternal.Deaths)),
    Live.Births = as.numeric(gsub(",", "", Live.Births)),
    # Creating proper date from Month.Ending.Date
    Date = mdy(Month.Ending.Date)
  ) %>%
  # Filter Total group and subgroup to avoid duplicates
  filter(Group == "Total" & Subgroup == "Total") %>%
  # Arranging by date
  arrange(Date)

head(VSRR_MM_clean)

# Structure of cleaned data
str(VSRR_MM_clean)
```

```{r}
# Trend plot to verify data
ggplot(VSRR_MM_clean, aes(x = Date, y = Maternal.Mortality.Rate)) +
  geom_line() + geom_point() + theme_minimal() + 
  labs(title = "US Maternal Mortality Rate Over Time", x = "Date", 
       y = "Deaths per 100,000 live births")


```

```{r}
# Calculating basic statistics
summary_stats <- VSRR_MM_clean %>%
  summarise( avg_rate = mean(Maternal.Mortality.Rate, na.rm = TRUE),
             min_rate = min(Maternal.Mortality.Rate, na.rm = TRUE),
             max_rate = max(Maternal.Mortality.Rate, na.rm = TRUE),
             start_rate = first(Maternal.Mortality.Rate),
             end_rate = last(Maternal.Mortality.Rate))

summary_stats
```

```{r}
# Time period labels
VSRR_MM_analysis <- VSRR_MM_clean %>%
  mutate(
    period = case_when(
      Date < as.Date("2020-03-01") ~ "Pre-COVID",
      Date >= as.Date("2020-03-01") & Date < as.Date("2022-06-24") ~ "COVID Era",
      Date >= as.Date("2022-06-24") ~ "Post-Dobbs",
      TRUE ~ "Other"
    )
  )

head(VSRR_MM_analysis)
```

```{r}
# Trend plot with periods
ggplot(VSRR_MM_analysis, aes(x = Date, y = Maternal.Mortality.Rate)) +
  geom_line(size = 1, color = "blue") +
  geom_point(size = 2, color = "blue") +
  # Period shading
  annotate("rect", 
           xmin = as.Date("2020-03-01"), 
           xmax = as.Date("2022-06-24"),
           ymin = -Inf, ymax = Inf,
           alpha = 0.2, fill = "red") +
  annotate("text", x = as.Date("2021-01-01"), y = 15,
           label = "COVID-19 Era\nHighest mortality 
           rates\nPeak: 33.8 deaths per 100,000",
           size = 3) +
  # Trend highlights
  geom_hline(yintercept = 18.9, linetype = "dashed", color = "gray50") +
  annotate("text", x = as.Date("2019-06-01"), y = 19.5,
           label = "Pre-COVID Average: 18.9", size = 3) +
  theme_minimal() +
  labs(
    title = "U.S. Maternal Mortality Rate Shows Dramatic COVID-19 Impact",
    subtitle = "Deaths per 100,000 live births increased 
    significantly during pandemic",
    x = "Year",
    y = "Deaths per 100,000 Live Births",
    caption = "Source: CDC National Center for Health Statistics\nNote: 
    Shaded area represents COVID-19 period"
  )
```

```{r}
# Statistics by period
period_stats <- VSRR_MM_analysis %>% group_by(period) %>% 
  summarise(avg_rate = mean(Maternal.Mortality.Rate, na.rm = TRUE), 
            min_rate = min(Maternal.Mortality.Rate, na.rm = TRUE),
            max_rate = max(Maternal.Mortality.Rate, na.rm = TRUE),
            num_months = n()) %>% arrange(min_rate)

# Percent changes between periods
changes <- VSRR_MM_analysis %>% group_by(period) %>% 
  summarise(start_rate = first(Maternal.Mortality.Rate),
            end_rate = last(Maternal.Mortality.Rate),
            total_change = end_rate - start_rate,
            percent_change = ((end_rate - start_rate) / start_rate) * 100)


# Statistics by Period
period_stats

# Changes within Each Period
changes

# Peak analysis
peak_analysis <- VSRR_MM_analysis %>% 
  filter(Maternal.Mortality.Rate == max(Maternal.Mortality.Rate)) %>% 
  select(Date, Maternal.Mortality.Rate, period)

#Peak Mortality Rate
peak_analysis
```

```{r}
# More detailed trend analysis
ggplot(VSRR_MM_analysis, aes(x = Date, y = Maternal.Mortality.Rate)) +
  # Trend lines for each period
  geom_smooth(aes(color = period), method = "lm", se = FALSE) + 
  geom_line(size = 1) + geom_point(aes(color = period), size = 2) +
  # Period-specific average lines
  geom_hline(data = period_stats, 
             aes(yintercept = avg_rate, color = period),
             linetype = "dashed", alpha = 0.5) +
  # Peak point Highlighted
  geom_point(data = peak_analysis, 
             aes(x = Date, y = Maternal.Mortality.Rate),
             color = "red", size = 4) +
  
  # Vertical lines for key events
  geom_vline(xintercept = as.Date("2020-03-01"),
             linetype = "dashed", color = "gray4") +
  geom_vline(xintercept = as.Date("2022-06-24"),
             linetype = "dashed", color = "gray4") + theme_minimal() +
  theme(legend.position = "bottom", 
        plot.title = element_text(size = 14, face = "bold"), 
        plot.subtitle = element_text(size = 12), 
        axis.title = element_text(size = 11)) + 
  labs(title = "US Maternal Mortality Rate: Three Distinct Periods",
    subtitle = paste("Peak of", 
                     round(max
                           (VSRR_MM_analysis$Maternal.Mortality.Rate, na.rm = 
                               TRUE), 1), 
                    "deaths per 100,000 live births reached during COVID Era"),
    x = "Date", y = "Deaths per 100,000 live births", color = "Period", 
    caption = "Source: CDC National Center for Health Statistics")
```

```{r}
# Yearly averages and changes
yearly_stats <- VSRR_MM_analysis %>% mutate(year = year(Date)) %>%
  group_by(year) %>% summarise(avg_rate = mean(Maternal.Mortality.Rate, 
                                               na.rm = TRUE), 
                               min_rate = min(Maternal.Mortality.Rate, 
                                              na.rm = TRUE),
    max_rate = max(Maternal.Mortality.Rate, na.rm = TRUE)
  ) %>% mutate(year_over_year_change = avg_rate - lag(avg_rate), 
               percent_change = (avg_rate - lag(avg_rate))/lag(avg_rate) * 100)

yearly_stats

# Key statistics for each period
period_analysis <- VSRR_MM_analysis %>% group_by(period) %>% 
  summarise(duration_months = n(), 
            avg_deaths = mean(Maternal.Deaths, na.rm = TRUE), 
            avg_births = mean(Live.Births, na.rm = TRUE), 
            peak_rate = max(Maternal.Mortality.Rate, na.rm = TRUE), 
            lowest_rate = min(Maternal.Mortality.Rate, na.rm = TRUE), 
            rate_volatility = sd(Maternal.Mortality.Rate, na.rm = TRUE)) %>% 
  mutate(deaths_per_month = avg_deaths/duration_months, 
         births_per_month = avg_births/duration_months)

period_analysis

```

```{r}
VSRR_MM_race <- VSRR_MM %>%
  # Filter for Race and Hispanic origin group
  filter(`Group` == "Race and Hispanic origin") %>%
  # Remove any rows with suppressed or missing data
  filter(!is.na(Maternal.Mortality.Rate)) %>%
  mutate(
    # Clean numeric columns
    Maternal.Deaths = as.numeric(gsub(",", "", Maternal.Deaths)),
    Live.Births = as.numeric(gsub(",", "", Live.Births)),
    Date = mdy(Month.Ending.Date),
    # Add period labels
    period = case_when(
      Date < as.Date("2020-03-01") ~ "Pre-COVID",
      Date >= as.Date("2020-03-01") & Date < as.Date("2022-06-24") ~ "COVID Era",
      Date >= as.Date("2022-06-24") ~ "Post-Dobbs",
      TRUE ~ "Other"
    )
  )

head(VSRR_MM_race)
```

```{r}
# Statistics by period and race/ethnicity
racial_stats <- VSRR_MM_race %>%
  group_by(period, Subgroup) %>%
  summarise(
    mean_rate = mean(Maternal.Mortality.Rate, na.rm = TRUE),
    median_rate = median(Maternal.Mortality.Rate, na.rm = TRUE),
    min_rate = min(Maternal.Mortality.Rate, na.rm = TRUE),
    max_rate = max(Maternal.Mortality.Rate, na.rm = TRUE),
    sd_rate = sd(Maternal.Mortality.Rate, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  arrange(period, desc(mean_rate))

# Disparities using reframe
disparity_calc <- VSRR_MM_race %>%
  group_by(Date, period) %>%
  reframe(
    Subgroup = Subgroup,
    Mortality_Rate = Maternal.Mortality.Rate,
    White_Rate = Maternal.Mortality.Rate[Subgroup == "White, Non-Hispanic"],
    Relative_Disparity = Maternal.Mortality.Rate / 
      Maternal.Mortality.Rate[Subgroup == "White, Non-Hispanic"]
  ) %>%
  filter(Subgroup != "White, Non-Hispanic") # Remove reference group

head(racial_stats)
head(disparity_calc)
```

```{r}
ggplot(VSRR_MM_race, aes(x = Date, y = Maternal.Mortality.Rate,
color = Subgroup)) + geom_line(linewidth = 1) +
geom_point(size = 2, alpha = 0.6) +
geom_vline(xintercept = as.Date("2020-03-01"),
linetype = "dashed", color = "gray4", alpha = 0.5) +
geom_vline(xintercept = as.Date("2022-06-24"),
linetype = "dashed", color = "gray4", alpha = 0.5) +
annotate("text", x = as.Date("2020-01-15"),
y = max(VSRR_MM_race$Maternal.Mortality.Rate, na.rm = TRUE),
label = "COVID-19 Start", angle = 90, vjust = -0.5) +
annotate("text", x = as.Date("2022-05-15"),
y = max(VSRR_MM_race$Maternal.Mortality.Rate, na.rm = TRUE),
label = "Dobbs Decision", angle = 90, vjust = -0.5) +
theme_minimal() + theme(legend.position = "right", plot.title =
element_text(size = 14, face = "bold"),
plot.subtitle = element_text(size = 12),
axis.text.x = element_text(angle = 45, hjust = 1)) +
labs(
title = "Maternal Mortality Rates by Race and Ethnicity",
subtitle = "Trends across COVID-19 and Dobbs decision periods (2019-2024)",
x = "Date",
y = "Deaths per 100,000 live births",
color = "Race/Ethnicity"
) 


```

```{r}
# Statistics by period and race/ethnicity
disparity_summary <- disparity_calc %>%
  group_by(period, Subgroup) %>%
  summarise(
    mean_disparity = mean(Relative_Disparity, na.rm = TRUE),
    median_disparity = median(Relative_Disparity, na.rm = TRUE),
    min_disparity = min(Relative_Disparity, na.rm = TRUE),
    max_disparity = max(Relative_Disparity, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  arrange(period, desc(mean_disparity))

head(disparity_summary)

# Period comparison with proper grouping
period_comparison <- racial_stats %>%
  select(period, Subgroup, mean_rate, max_rate) %>%
  pivot_wider(
    id_cols = Subgroup,
    names_from = period,
    values_from = c(mean_rate, max_rate)
  )

head(period_comparison)
```



